Improving the portuguese hyphenation rules

### 2023-11-29 article/article.tex
fix issue #42
fix #39

For any disagreement, we must be cautious and examine each case.
We might be tempted to chose the majority vote approach, selection
those hyphenations that have the most supporters, but it is unclear 
how the hyphenations on those online dictionaries were curated, 
many might use algorithmic approach, leading to potential flaws shared
among them. 
For example, 5 dictionaries use the hyphenation 
\emph{quart-zo} while just one uses the rightful one, \emph{quar-tzo}.
The same happens with \emph{de-sem-ba-i-nhar} and \emph{de-sem-bai-nhar},
the former one receiving the majority of the votes, despite being wrong.
Some strings might accept more than one hyphenation. For example, 
\emph{sub-li-nha} and \emph{su-bli-nha} represent two different words.
The former is the substantive \emph{underline} and the latter is a flexed form of the verb \emph{to underline}.
\todo{Aline, segue a lista dos casos duvidosos que encontrei com votos 5 a 1. Marquei em negrito aqueles
    que considero ser a forma correta.
    quart-zo(5), \textbf{quar-tzo}(1) (assim também trocar quart-zi-to por \textbf{quar-tzi-to)}; \textbf{sub-li-nha}(5), \textbf{su-bli-nha}(1); 
    de-sem-ba-i-nhar(5), \textbf{de-sem-bai-nhar}(1); ap-nei-a(5), \textbf{a-pnei-a}(1); 
    disp-nei-a(5), dispneia(1) (\textbf{dis-pnei-a}, forma não encontrada nos dicionários); 
    hi-pe-ral-ge-si-a(5), \textbf{hi-per-al-ge-si-a}(1);
    in-de-is-cen-te(5), \textbf{in-deis-cen-te}(1); jo-in-vi-len-se(5), \textbf{join-vi-len-se}(1); 
    ne-er-lan-dês(5), \textbf{neer-lan-dês}(1). Os demais casos de 5 a 1 são proparoxítonas aparentes.
}
The procedure to create the second list of hyphenated words was:
\begin{enumerate*}[label=\arabic*)]
  \item select those hyphenations agreed by five dictionaries that have no other hyphenation proposal (5 to 0 votes);
  \item from those that have another proposal, check if it is an apparent proparoxytone,
      \begin{enumerate*}[label=\alph*)]
	  \item if it has a diacritic, select the paroxytone option,
	  \item if not, select the proparoxytone option.
      \end{enumerate*}
  \item if it is not an apparent proparoxytone case, select the hyphenation with five votes.
\end{enumerate*}
This procedure is implemented in the script \texttt{get51.sh}.
Just a few hyphenations were manually corrected. They are: \emph{quar-tzo}, 
\emph{quar-tzi-to}, \emph{su-bli-nha} (was added), \emph{de-sem-bai-nhar},
\emph{a-pnei-a}, \emph{dis-pnei-a}, \emph{hi-per-al-ge-si-a}, \emph{join-vi-len-se}, and \emph{neer-lan-dês}.
The final list is stored in \texttt{hyphenations5.dic} and holds \num{\NumberOfFiveAgrees} additional hyphenations.
Using the two lists created so far, we have in total \num{\NumberOfSixFiveAgrees} hyphenations.

@@

	\begin{table}
	\scriptsize
	\caption{Hyphenation in the dictionaries of words having conflict between morphological and syllabic information.}
	\label{tab-mor-syl}
        \begin{tabular}{ll*{6}{c}}
	\toprule
	    word                   & hyphenation  & \multicolumn{6}{c}{dictionary} \\
             &                                    & michaelis & priberam & wikitionary & aulete & portal & dicio \\
	\midrule
        \multirow{2}{*}{sublinhar} & su-bli-nhar  &           & x        & x           & x      &        &       \\
				   & sub-li-nhar  & x         &          &             &        & x      &       \\
	\multirow{2}{*}{reiniciar} & rei-ni-ci-ar &           &          & x           &        & x      &       \\
				   & re-i-ni-ci-ar & x        & x        &             & x      &        & x     \\
	\bottomrule
	\end{tabular}
	\end{table}

>>>>

For any disagreement, we must be cautious and examine each case.
We might be tempted to chose the majority vote approach, selection
those hyphenations that have the most supporters, but it is unclear 
how the hyphenations on those online dictionaries were curated, 
many might use algorithmic approach, leading to potential flaws shared
among them. 
For example, 5 dictionaries use the hyphenation 
\emph{quart-zo} while just one uses the rightful one, \emph{quar-tzo}.
The same happens with \emph{de-sem-ba-i-nhar} and \emph{de-sem-bai-nhar},
the former one receiving the majority of the votes, despite being wrong.
Some strings might accept more than one hyphenation. For example, 
\emph{sub-li-nha} and \emph{su-bli-nha} represent two different words.
The former is the substantive \emph{underline} and the latter is a flexed form of the verb \emph{to underline}.
\todo{Aline, segue a lista dos casos duvidosos que encontrei com votos 5 a 1. Marquei em negrito aqueles
    que considero ser a forma correta.
    quart-zo(5), \textbf{quar-tzo}(1) (assim também trocar quart-zi-to por \textbf{quar-tzi-to)}; \textbf{sub-li-nha}(5), \textbf{su-bli-nha}(1); 
    de-sem-ba-i-nhar(5), \textbf{de-sem-bai-nhar}(1); ap-nei-a(5), \textbf{a-pnei-a}(1); 
    disp-nei-a(5), dispneia(1) (\textbf{dis-pnei-a}, forma não encontrada nos dicionários); 
    hi-pe-ral-ge-si-a(5), \textbf{hi-per-al-ge-si-a}(1);
    in-de-is-cen-te(5), \textbf{in-deis-cen-te}(1); jo-in-vi-len-se(5), \textbf{join-vi-len-se}(1); 
    ne-er-lan-dês(5), \textbf{neer-lan-dês}(1). Os demais casos de 5 a 1 são proparoxítonas aparentes.
}
The procedure to create the second list of hyphenated words was:
\begin{enumerate*}[label=\arabic*)]
  \item select those hyphenations agreed by five dictionaries that have no other hyphenation proposal (5 to 0 votes);
  \item from those that have another proposal, check if it is an apparent proparoxytone,
      \begin{enumerate*}[label=\alph*)]
	  \item if it has a diacritic, select the paroxytone option (e.g. select \emph{po-lí-cia} instead of \emph{po-lí-ci-a} and \emph{pe-rí-neo} instead of \emph{pe-rí-ne-o}),
	  \item if not, select the proparoxytone option (e.g. select \emph{au-to-ri-a} instead of \emph{au-to-ri-a} and \emph{mar-ce-nei-ro} instead of \emph{mar-ce-neiro})\footnote{We just have to make sure we are not selecting those cases where there is a sequence of the form \texttt{[gq]u-[aeio]} (using regex notation), as in \emph{a-pa-zi-gu-ar} and \emph{en-xa-gu-ar}.}.
	      % [gq]u-[aeio]
      \end{enumerate*}
  \item if it is not an apparent proparoxytone case, select the hyphenation with five votes.
\end{enumerate*}
This procedure is implemented in the script \texttt{get51.sh}.
Just a few hyphenations were manually corrected. They are: \emph{quar-tzo}, 
\emph{quar-tzi-to}, \emph{su-bli-nha} (was added), \emph{de-sem-bai-nhar},
\emph{a-pnei-a}, \emph{dis-pnei-a}, \emph{hi-per-al-ge-si-a}, \emph{join-vi-len-se}, and \emph{neer-lan-dês}.
The final list is stored in \texttt{hyphenations5.dic} and holds \num{\NumberOfFiveAgrees} additional hyphenations.
Using the two lists created so far, we have in total \num{\NumberOfSixFiveAgrees} hyphenations.

@@

	\begin{table}
	\scriptsize
	\caption{Hyphenation in the dictionaries of words having conflict between morphological and syllabic information.}
	\label{tab-mor-syl}
        \begin{tabular}{ll*{6}{c}}
	\toprule
	    word                   & hyphenation  & \multicolumn{6}{c}{dictionary} \\
             &                                    & Michaelis & Priberam & Wikitionary & Aulete & Portal & Dicio \\
	\midrule
        \multirow{2}{*}{sublinhar} & su-bli-nhar  &           & x        & x           & x      &        &       \\
				   & sub-li-nhar  & x         &          &             &        & x      &       \\
	\multirow{2}{*}{reiniciar} & rei-ni-ci-ar &           &          & x           &        & x      &       \\
				   & re-i-ni-ci-ar & x        & x        &             & x      &        & x     \\
	\bottomrule
	\end{tabular}
	\end{table}
